---
title: "SQL handbook"
author: "Tony Duan"
execute:
  warning: false
  error: false
format:
  html:
    toc: true
    toc-location: right
    code-fold: show
    code-tools: true
    number-sections: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---


This document provides a comprehensive guide to performing common data manipulation tasks using SQL, R, and Python. It serves as a reference for understanding how to achieve similar outcomes across these three popular data analysis tools.

## Introduction to Data Analytics Tools

In data analysis, we work with different tools, each with its strengths:

- **SQL**: Excellent for structured query and database operations
- **R**: Powerful statistical computing and data visualization
- **Python**: Versatile programming with great data science libraries

This tutorial shows how to accomplish the same data manipulation tasks across all three tools, helping you choose the right approach for your needs.


# Setup and Configuration

```{r}
#| code-fold: true
#| output: false
library(reticulate)
py_require(c("pandas","Great-Tables","polars","pyarrow"))

# Load libraries for database interaction, data manipulation, and connections.
library(DBI)
library(tidyverse)
library(RSQLite)
library(connections)
library(duckdb)

# Prepare the R dataframes.
# Remove the existing mtcars dataset if it exists.
rm(mtcars)
# Create the mtcars dataframe from the base R dataset, adding the row names as a new column.
mtcars = cbind(model_name = rownames(mtcars), mtcars) |> head(10)
# Create the iris dataframe from the base R dataset.
iris = iris |> head(10)

# Remove the database file if it already exists to start with a clean slate.
if (file.exists("my-db.duckdb")) {
  file.remove("my-db.duckdb")
}
# Establish a connection to the DuckDB database.
con <- dbConnect(duckdb(), dbdir = "my-db.duckdb", read_only = FALSE)
# Write the iris and mtcars dataframes to the database as tables.
dbWriteTable(con, "iris_table", iris, overwrite = TRUE)
dbWriteTable(con, "mtcars_table", mtcars, overwrite = TRUE)

```


```{python}
#| code-fold: true
#| output: false
import pandas as pd
import polars as pl
import os

from platform import python_version
#print(python_version())

cities_pd = pd.DataFrame({
    'country': ['NL', 'NL', 'NL', 'US', 'US', 'US', 'US', 'US', 'US'],
    'name': ['Amsterdam', 'Amsterdam', 'Amsterdam', 'Seattle', 'Seattle', 'Seattle', 'New York City', 'New York City', 'New York City'],
    'year': [2000, 2010, 2020, 2000, 2010, 2020, 2000, 2010, 2020],
    'population': [1005, 1065, 1158, 564, 608, 738, 8015, 8175, 8772]
})

# Make the R dataframes available in the Python environment.
mtcars_pd = r.mtcars
iris_pd = r.iris
mtcars_pl = pl.from_pandas(mtcars_pd)
iris_pl = pl.from_pandas(iris_pd)
cities_pl=pl.from_pandas(cities_pd)
```


- There are tables `iris_table` and `mtcars_table` in the DuckDB database `my-db.duckdb`.
- There are dataframes `iris` and `mtcars` in the R environment.
- There are dataframes `iris` and `mtcars` in the Python environment.

# Getting Started with Data Exploration

## Show All Tables

This section demonstrates how to list all available tables or dataframes in each environment.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Show all tables in the connected database.
SHOW ALL TABLES;
```

## R

```{r}
# List all dataframes in the current R environment.
dflist <- Filter(is.data.frame, as.list(.GlobalEnv))
names(dflist)
```

## Python pandas

```{python}
# List all pandas dataframes in the current Python environment.
import pandas as pd
alldfs = [var for var in dir() if isinstance(eval(var), pd.core.frame.DataFrame)]
print(alldfs)
```

## Python Polars

```{python}
import polars as pl

alldfs = [name for name, val in globals().items() if isinstance(val, pl.DataFrame)]
print(alldfs)
```


:::

# Describe a Table

This section shows how to get a summary of a table's structure and statistics.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Describe the columns and data types of the mtcars_table.
DESCRIBE mtcars_table;
```

## R

```{r}
# Provide a detailed summary of the mtcars dataframe.
skimr::skim(mtcars)
```

## Python pandas

```{python}
# Generate descriptive statistics for the mtcars dataframe.
mtcars_pd.describe(include='all')
```



## Python Polars

```{python}
mtcars_pl.describe()
```




:::

# Show Column Names

This section demonstrates how to retrieve the column names of a table.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Show information about the columns in mtcars_table.
PRAGMA table_info(mtcars_table);
```

## R

```{r}
# Get the names of the columns in the mtcars dataframe.
names(mtcars)
```

## Python pandas

```{python}
# Get a list of column names from the mtcars dataframe.
import pandas as pd
list(mtcars_pd.columns.values)
```


## Python Polars

```{python}
# Get a list of column names from the mtcars dataframe.
mtcars_pl.columns
```



:::

# Select

## Select Top 6 and Rename

This section shows how to select a subset of columns, rename them, and limit the number of rows returned.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Select and rename columns, limiting the result to the top 6 rows.
SELECT model_name as model, mpg, cyl FROM mtcars_table LIMIT 6;
```

## R

```{r}
# Select and rename columns from the first 6 rows of the mtcars dataframe.
head(mtcars, 6) |> select(model = model_name, mpg, cyl)
```

## Python pandas

```{python}
# Rename a column in the mtcars dataframe.
mtcars_pd.rename(columns={'model_name': 'model'})
```

## Python Polars

```{python}
mtcars_pl.select(pl.col("model_name").alias("model"), pl.col("mpg"), pl.col("cyl")).head(6)
```

:::

## Select Distinct

This section demonstrates how to retrieve unique rows based on specified columns.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Select distinct combinations of mpg and cyl.
SELECT DISTINCT mpg, cyl FROM mtcars_table;
```

## R

```{r}
# Get distinct rows based on mpg and cyl.
mtcars |> distinct(mpg, cyl)
```

## Python pandas

```{python}
# Select specific columns and drop duplicate rows.
df = mtcars_pd[["mpg", "cyl"]]
print(df.drop_duplicates())
```

## Python Polars

```{python}
mtcars_pl.select(pl.col("mpg"), pl.col("cyl")).unique()
```

:::

# Check Row and Column Number

This section shows how to find the dimensions of a table.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Get the total number of rows.
SELECT count(*) AS row_number FROM mtcars_table;
```

```{sql}
#| connection: con
-- Get the total number of columns.
SELECT count(*) AS column_number FROM (DESCRIBE mtcars_table);
```

## R

```{r}
# Get the number of rows.
nrow(mtcars)
```

```{r}
# Get the number of columns.
ncol(mtcars)
```

## Python pandas

```{python}
# Get the number of rows.
mtcars_pd.shape[0]
```

```{python}
# Get the number of columns.
mtcars_pd.shape[1]
```



## Python Polars

```{python}
# Get the number of rows.
mtcars_pl.shape[0]
```

```{python}
# Get the number of columns.
mtcars_pl.shape[1]
```



:::

# Create Column

This section demonstrates how to add a new column to a table based on existing data.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Create a new column by performing a calculation on an existing column.
SELECT *, mpg + 1 AS new_mpg FROM mtcars_table;
```

## R

```{r}
# Add a new column to the dataframe.
mtcars |> mutate(new_mpg = mpg + 1)
```

## Python pandas

```{python}
# Create a new column in the dataframe.
mtcars_pd["new_mpg"] = mtcars_pd["mpg"] + 1
mtcars_pd
```

## Python Polars

```{python}
mtcars_pl.with_columns((pl.col("mpg") + 1).alias("new_mpg"))
```

:::

# Filter

This section shows how to select rows that meet specific criteria.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Use the AND operator to filter rows based on multiple conditions.
SELECT * FROM mtcars_table WHERE mpg = 21 AND cyl = 6;
```

```{sql}
#| connection: con
-- Use the OR operator to filter rows.
SELECT * FROM mtcars_table WHERE mpg = 21 OR cyl = 6;
```

## R

```{r}
# Use the & operator for "and" filtering.
mtcars |> filter(mpg == 21 & cyl == 6)
```

```{r}
# Use the | operator for "or" filtering.
mtcars |> filter(mpg == 21 | cyl == 6)
```

## Python pandas

```{python}
# Use the query method for "and" filtering.
mtcars_pd.query('mpg == 21 and cyl == 6')
```

```{python}
# Use the query method for "or" filtering.
mtcars_pd.query('mpg == 21 or cyl == 6')
```

## Python Polars

```{python}
# "and" filtering with polars
mtcars_pl.filter((pl.col("mpg") == 21) & (pl.col("cyl") == 6))
```

```{python}
# "or" filtering with polars
mtcars_pl.filter((pl.col("mpg") == 21) | (pl.col("cyl") == 6))
```

:::

# Order

This section demonstrates how to sort a table based on one or more columns.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Order the results by mpg in descending order and show the top 3.
SELECT model_name AS model, mpg, cyl FROM mtcars_table ORDER BY mpg DESC LIMIT 3;
```

## R

```{r}
# Arrange the dataframe by mpg in descending order.
mtcars |> select(model = model_name, mpg, cyl) |> arrange(desc(mpg)) |> head(3)
```

## Python pandas

```{python}
# Sort the dataframe by mpg in descending order.
mtcars_pd[["model_name", "mpg", "cyl"]].sort_values(by='mpg', ascending=False).head(3)
```

## Python Polars

```{python}
mtcars_pl.select("model_name", "mpg", "cyl").sort("mpg", descending=True).head(3)
```

:::

# Group By

This section shows how to group rows and perform aggregate calculations.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Group by model_name and calculate the sum of mpg and the mean of cyl.
SELECT model_name, sum(mpg) AS total_mpg, mean(cyl) AS cyl_mean FROM mtcars_table
GROUP BY 1 LIMIT 5;
```

## R

```{r}
# Group by model_name and summarize the data.
mtcars |> group_by(model_name) |> summarise(total_mpg = sum(mpg), cyl_mean = mean(cyl)) |> head(5)
```

## Python pandas

```{python}
# Group by model_name and aggregate the data.
mtcars_pd.groupby('model_name').agg({'mpg': 'sum', 'cyl': 'mean'}).head(5)
```

## Python Polars

```{python}
mtcars_pl.group_by("model_name").agg([
    pl.sum("mpg").alias("total_mpg"),
    pl.mean("cyl").alias("cyl_mean")
]).head(5)
```

:::

# Create Table

## CREATE OR REPLACE

This command creates a new table or overwrites an existing one.

```{sql}
#| connection: con
-- Create a temporary table, replacing it if it already exists.
CREATE OR REPLACE TEMP TABLE mtcars_table_group AS
SELECT model_name, sum(mpg) AS total_mpg, mean(cyl) AS cyl_mean FROM mtcars_table
GROUP BY 1;
```

## CREATE TABLE IF NOT EXISTS

This command creates a table only if it does not already exist.

```{sql}
#| connection: con
-- Create a new table only if it does not already exist.
CREATE TABLE IF NOT EXISTS new_mtcars_table_group AS
SELECT model_name, sum(mpg) AS total_mpg, mean(cyl) AS cyl_mean FROM mtcars_table
GROUP BY 1;
```

# Unique

## Check Unique

This section demonstrates how to verify the uniqueness of values in a column.

::: panel-tabset

### SQL

```{sql}
#| connection: con
-- Count the total number of rows and the number of distinct model names.
SELECT count(*), count(DISTINCT model_name) FROM mtcars_table;
```

### Python pandas

```{python}
# Count the total number of rows and the number of distinct model names.
print(len(mtcars_pd), mtcars_pd.model_name.nunique())
```

### Python Polars

```{python}
# Count the total number of rows and the number of distinct model names.
mtcars_pl.select(pl.count(), pl.col("model_name").n_unique())
```

:::

## Get Duplicate and Non-Duplicate Data

This section demonstrates how to get duplicate and non-duplicate data based on a single column.

::: panel-tabset

### SQL

#### Show all Duplicate

```{sql}
#| connection: con
-- Get duplicate rows based on the mpg column.
SELECT * FROM mtcars_table WHERE mpg IN (SELECT mpg FROM mtcars_table GROUP BY mpg HAVING count(*) > 1);
```

#### Keep non-Duplicate

```{sql}
#| connection: con
-- Get non-duplicate rows based on the mpg column.
SELECT * FROM mtcars_table WHERE mpg IN (SELECT mpg FROM mtcars_table GROUP BY mpg HAVING count(*) = 1);
```

### R

#### Show all Duplicate


```{r}
# Get duplicate rows based on the mpg column.
mtcars |> filter(duplicated(mpg) | duplicated(mpg, fromLast = TRUE))
```

#### keep non Duplicate


```{r}
# Get non-duplicate rows based on the mpg column.
mtcars |> filter(!duplicated(mpg) & !duplicated(mpg, fromLast = TRUE))
```

### Python pandas

#### Show all Duplicate


```{python}
# Get duplicate rows based on the mpg column.
mtcars_pd[mtcars_pd.duplicated(subset=['mpg'], keep=False)]
```
#### keep non Duplicate

```{python}
# Get non-duplicate rows based on the mpg column.
mtcars_pd[~mtcars_pd.duplicated(subset=['mpg'], keep=False)]
```

### Python Polars

```{python}
# Get duplicate rows based on the mpg column.
mtcars_pl.filter(pl.col("mpg").is_duplicated())
```

#### keep non Duplicate

```{python}
# Get non-duplicate rows based on the mpg column.
mtcars_pl.filter(pl.col("mpg").is_unique())
```

:::


# Join

```{sql}
#| connection: con
-- Select all data from the newly created temporary table.
SELECT * FROM mtcars_table_group t1;
```

## Left Join

This section shows how to perform a left join to combine data from two tables.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Perform a left join between mtcars_table and mtcars_table_group.
SELECT t1.model_name, t1.mpg, t1.cyl, t2.* FROM mtcars_table t1
LEFT JOIN mtcars_table_group t2 ON t1.model_name = t2.model_name
LIMIT 5;
```

## R

```{r}
# Perform a left join on the mtcars dataframe with itself.
mtcars |> left_join(mtcars, by = join_by(model_name == model_name)) |> head(5)
```

## Python pandas

```{python}
# Perform a left join using pandas.
pd.merge(mtcars_pd, mtcars_pd, left_on='model_name', right_on='model_name', how='left').head(5)
```

## Python Polars

```{python}
mtcars_pl.join(mtcars_pl, on="model_name", how="left").head(5)
```

:::

## Inner Join

This section demonstrates how to perform an inner join.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Perform an inner join with a subquery.
SELECT t1.model_name, t1.mpg, t1.cyl, t2.* FROM mtcars_table t1
INNER JOIN (SELECT * FROM mtcars_table_group LIMIT 5) t2 ON t1.model_name = t2.model_name;
```

## R

```{r}
# Perform an inner join.
mtcars |> inner_join(mtcars, by = join_by(model_name == model_name)) |> head(5)
```

## Python pandas

```{python}
# Perform an inner join with pandas.
pd.merge(mtcars_pd, mtcars_pd, left_on='model_name', right_on='model_name', how='inner').head(5)
```

## Python Polars

```{python}
mtcars_pl.join(mtcars_pl, on="model_name", how="inner").head(5)
```

:::

# Append Rows

## Append Without Duplicate Elimination (`union all`)

This section shows how to combine rows from two tables, keeping all duplicates.

::: panel-tabset

### SQL

```{sql}
#| connection: con
-- Count the rows in the original table.
SELECT count(*) FROM mtcars_table;
```

```{sql}
#| connection: con
-- Create a new table by appending mtcars_table to itself, including duplicates.
CREATE TEMP TABLE double_mtcars_table AS
SELECT * FROM mtcars_table
UNION ALL
SELECT * FROM mtcars_table;
```

```{sql}
#| connection: con
-- Count the rows in the new table.
SELECT count(*) FROM double_mtcars_table;
```

### R

```{r}
# Append rows using bind_rows.
mtcars %>% bind_rows(mtcars)
```

### Python pandas

```{python}
# Concatenate dataframes, keeping all rows.
pd.concat([mtcars_pd, mtcars_pd], ignore_index=True)
```

:::

## Append With Duplicate Elimination (`union`)

This section demonstrates how to combine rows while removing duplicate entries.

::: panel-tabset

### SQL

```{sql}
#| connection: con
-- Create a table, removing duplicate rows.
CREATE OR REPLACE TEMP TABLE union_mtcars_table AS
SELECT * FROM mtcars_table
UNION
SELECT * FROM mtcars_table;
```

```{sql}
#| connection: con
-- Count the rows in the new table (should be same as original since all rows are identical).
SELECT count(*) AS union_row_count FROM union_mtcars_table;
```

### R

```{r}
# Append rows and then remove duplicates.
mtcars %>% bind_rows(mtcars) |> distinct()
```

### Python pandas

```{python}
# Concatenate dataframes and then drop duplicates.
pd.concat([mtcars_pd, mtcars_pd], ignore_index=True).drop_duplicates().reset_index(drop=True)
```

:::

# Delete Rows

This section shows how to remove rows from a table based on a condition.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Delete a specific row from the table.
DELETE FROM mtcars_table WHERE model_name = 'Mazda RX4';
```

```{sql}
#| connection: con
-- Verify the row count after deletion.
SELECT count(*) FROM mtcars_table;
```

## R

```{r}
# Filter out the specified row.
mtcars |> filter(model_name != 'Mazda RX4')
```

## Python pandas

```{python}
# Select all rows except the one specified.
mtcars_pd[mtcars_pd['model_name'] != 'Mazda RX4']
```

## Python Polars

```{python}
mtcars_pl.filter(pl.col("model_name") != 'Mazda RX4')
```

:::

# Update Rows

This section demonstrates how to modify existing data in a table.

::: panel-tabset

## SQL

```{sql}
#| connection: con
-- Select the row to be updated.
SELECT model_name, mpg FROM mtcars_table WHERE model_name = 'Mazda RX4 Wag';
```

```{sql}
#| connection: con
-- Update the value of a column for a specific row.
UPDATE mtcars_table
SET mpg = 999
WHERE model_name = 'Mazda RX4 Wag';
```

```{sql}
#| connection: con
-- Verify the update.
SELECT model_name, mpg FROM mtcars_table WHERE model_name = 'Mazda RX4 Wag';
```

## R

```{r}
# Update a value within a filtered selection.
mtcars |> filter(model_name == 'Mazda RX4 Wag') |> mutate(mpg = 999)
```

## Python pandas

```{python}
# Update a specific value using .loc.
mtcars_pd.loc[mtcars_pd['model_name'] == 'Mazda RX4 Wag', 'mpg'] = 999
mtcars_pd
```

## Python Polars

```{python}
mtcars_pl.with_columns(
    pl.when(pl.col("model_name") == 'Mazda RX4 Wag')
    .then(999)
    .otherwise(pl.col("mpg"))
    .alias("mpg")
)
```

:::

# Drop Table

This section shows how to remove a table from the database.

Before dropping the table:

```{sql}
#| connection: con
-- Show all tables before dropping one.
SHOW ALL TABLES;
```

Dropping the table:

```{sql}
#| connection: con
-- Drop the specified table if it exists.
DROP TABLE IF EXISTS mtcars_table_group;
```

After dropping the table:

```{sql}
#| connection: con
-- Show all tables again to confirm the deletion.
SHOW ALL TABLES;
```

# PIVOT

This section demonstrates how to transform data from a long to a wide format.

```{sql}
#| connection: con
-- Create a temporary table for the pivot demonstration.
DROP TABLE IF EXISTS cities;
CREATE TEMP TABLE cities (
    country VARCHAR, name VARCHAR, year INTEGER, population INTEGER
);
INSERT INTO cities VALUES
    ('NL', 'Amsterdam', 2000, 1005),
    ('NL', 'Amsterdam', 2010, 1065),
    ('NL', 'Amsterdam', 2020, 1158),
    ('US', 'Seattle', 2000, 564),
    ('US', 'Seattle', 2010, 608),
    ('US', 'Seattle', 2020, 738),
    ('US', 'New York City', 2000, 8015),
    ('US', 'New York City', 2010, 8175),
    ('US', 'New York City', 2020, 8772);
```

```{sql}
#| connection: con
-- View the raw data.
SELECT * FROM cities;
```

## PIVOT on One Column

::: panel-tabset

### SQL

```{sql}
#| connection: con
-- Pivot the table on the 'year' column.
PIVOT cities
ON year
USING sum(population)
GROUP BY country;
```

### Python pandas

```{python}
cities_pd.pivot_table(index='country', columns='year', values='population', aggfunc='sum').reset_index()
```

### Python Polars

```{python}
cities_pl.pivot(index='country', columns='year', values='population', aggregate_function='sum')
```

:::

## PIVOT on Two Columns

::: panel-tabset

### SQL

```{sql}
#| connection: con
-- Pivot on both 'country' and 'name'.
PIVOT cities
ON country, name
USING sum(population);
```

### Python pandas

```{python}
cities_pd.pivot_table(index=['country', 'name'], columns='year', values='population', aggfunc='sum').reset_index()
```

### Python Polars

```{python}
cities_pl.pivot(index=['country', 'name'], columns='year', values='population', aggregate_function='sum')
```

:::

# UNPIVOT

This section shows how to transform data from a wide to a long format.

::: panel-tabset

### SQL

```{sql}
#| connection: con
-- Unpivot the previously pivoted table.
UNPIVOT
(
  PIVOT cities
  ON year
  USING sum(population)
  GROUP BY country
)
ON COLUMNS(* EXCLUDE (country))
INTO
    NAME year
    VALUE population;
```

### Python pandas




```{python}
cities_pivot_pd = cities_pd.pivot_table(index='country', columns='year', values='population', aggfunc='sum').reset_index()
cities_pivot_pd.melt(id_vars='country', value_vars=[2000, 2010, 2020], var_name='year', value_name='population')
```

### Python Polars

```{python}
cities_pivot_pl = cities_pl.pivot(index='country', columns='year', values='population', aggregate_function='sum')
cities_pivot_pl.unpivot(index='country', on=['2000', '2010', '2020'], variable_name='year', value_name='population')
```

:::

# EXPLAIN

This section demonstrates how to analyze the execution plan of a query.

```{sql}
#| eval: false
#| connection: con
-- Show the logical execution plan of a query.
EXPLAIN
SELECT t1.model_name, t1.mpg, t1.cyl, t2.* FROM mtcars_table t1
INNER JOIN (SELECT * FROM mtcars_table LIMIT 5) t2 ON t1.model_name = t2.model_name;
```

```{sql}
#| eval: false
#| connection: con
-- Show the detailed execution plan, including timing information.
EXPLAIN ANALYZE
SELECT t1.model_name, t1.mpg, t1.cyl, t2.* FROM mtcars_table t1
INNER JOIN (SELECT * FROM mtcars_table LIMIT 5) t2 ON t1.model_name = t2.model_name;
```

```{r}
# Disconnect from the database and shut it down.
dbDisconnect(con, shutdown = TRUE)
```

# Display Beautiful Table

This section shows how to create nicely formatted tables for presentation.

::: panel-tabset

## R

```{r}
# Use the gt library to create a formatted table.
library(gt)
mtcars |> gt()
```

## Python pandas

```{python}
# Use the great_tables library to create a formatted table.
from great_tables import GT
GT(mtcars_pd)
```

## Python Polars

```{python}
# Use the great_tables library to create a formatted table.
from great_tables import GT
GT(mtcars_pl)
```




:::

# Reference:

- [DuckDB R Client Documentation](https://duckdb.org/docs/stable/clients/r.html)
- [DuckDB SQL Statements](https://duckdb.org/docs/stable/sql/statements)
